From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-09-21 WOLFHPC15/article/component.tex
corrections article reviewer2 + related + reduction taille -  Helene

\begin{description}
\item[Sequence component (SEQ)] It is the direct representation of a sequence node of $\Gamma_{tsp}$ or $\Gamma_{data}$. This component sequentially calls an ordered list of other components. It exposes an ordered \emph{use-multiple} port to be connected to the components to call in sequence.

\item[Parallel component (PAR)] It is the direct representation of a parallel node of $\Gamma_{tsp}$ (not $\Gamma_{data}$). This component simultaneously calls a set of other components. It exposes a \emph{use-multiple} port to be connected to the components to call in parallel.

\item[Synchronization component (SYNC)] It is the direct representation to an update leaf of $\Gamma_{tsp}$ or $\Gamma_{data}$. This component calls the synchronization of a given data. It exposes a \emph{use} port to be connected to the data to be updated.
\end{description}

>>>>

% \begin{description}
% \item[Sequence component (SEQ)] It is the direct representation of a sequence node of $\Gamma_{tsp}$ or $\Gamma_{data}$. This component sequentially calls an ordered list of other components. It exposes an ordered \emph{use-multiple} port to be connected to the components to call in sequence.

% \item[Parallel component (PAR)] It is the direct representation of a parallel node of $\Gamma_{tsp}$ (not $\Gamma_{data}$). This component simultaneously calls a set of other components. It exposes a \emph{use-multiple} port to be connected to the components to call in parallel.

% \item[Synchronization component (SYNC)] It is the direct representation to an update leaf of $\Gamma_{tsp}$ or $\Gamma_{data}$. This component calls the synchronization of a given data. It exposes a \emph{use} port to be connected to the data to be updated.
% \end{description}

\textbf{Sequence component (SEQ).} It is the direct representation of a sequence node of $\Gamma_{tsp}$ or $\Gamma_{data}$. This component sequentially calls an ordered list of other components. It exposes an ordered \emph{use-multiple} port to be connected to the components to call in sequence.

\textbf{Parallel component (PAR).} It is the direct representation of a parallel node of $\Gamma_{tsp}$ (not $\Gamma_{data}$). This component simultaneously calls a set of other components. It exposes a \emph{use-multiple} port to be connected to the components to call in parallel.

\textbf{Synchronization component (SYNC).} It is the direct representation to an update leaf of $\Gamma_{tsp}$ or $\Gamma_{data}$. This component calls the synchronization of a given data. It exposes a \emph{use} port to be connected to the data to be updated.

### 2015-09-21 WOLFHPC15/article/concept.tex
corrections article reviewer2 + related + reduction taille -  Helene

%To resume, concepts used in this paper to define a stencil kernel are \emph{mesh}, \emph{mesh entities}, \emph{data}, \emph{computation domain}, \emph{numerical expression} and \emph{stencil shape}. 
While stencil kernels have been studied a lot, the formalization of real overall mesh-based numerical simulations is a rather new contribution. Actually, paying attention to complex numerical simulations, it appears that most of them are composed of more than one stencil kernel, with one or more stencil shapes and of additional local computations. For example, how to formalize and parallelize a numerical simulation which chains the stencil kernels of Figures~\ref{fig:ex1} and~\ref{fig:ex2}. The next section formalizes concepts of \emph{stencil kernel}, \emph{local kernel} and \emph{multi-stencil program}.

@@

We define a multi-stencil program as a sextuplet:
\begin{equation} 
\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma),
\label{eq:msp}
\end{equation}
where $T$ is the set of time iteration to run the simulation, $\mathcal{M}$ is the mesh of the simulation, $\mathcal{E}$ is the set of mesh entities, $\mathcal{D}$ is the set of computation domains used for computations, $\Delta$ is the set of data of the simulation, each one mapped onto a mesh entity, and finally $\Gamma$ is the set of computations. The set of computations $\Gamma$ is an ordered list of stencil and local kernels.
One can notice that this work, for now, is limited to a single type of mesh in a given simulation.

>>>>

%To resume, concepts used in this paper to define a stencil kernel are \emph{mesh}, \emph{mesh entities}, \emph{data}, \emph{computation domain}, \emph{numerical expression} and \emph{stencil shape}. 
While stencil kernels have been studied a lot, the formalization of real overall mesh-based numerical simulations is poorly studied and not exactly for the same kind of appication~\cite{Ragan-Kelley:2013:HLC:2491956.2462176}. Actually, paying attention to complex numerical simulations, it appears that most of them are composed of more than one stencil kernel, with one or more stencil shapes and of additional local computations. For example, we would like to formalize and parallelize a numerical simulation which chains the stencil kernels of Figures~\ref{fig:ex1} and~\ref{fig:ex2}. The next section formalizes concepts of \emph{stencil kernel}, \emph{local kernel} and \emph{multi-stencil program}.

@@

We define a multi-stencil program as a sextuplet:
\begin{equation} 
\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma),
\label{eq:msp}
\end{equation}
where $T$ is the set of time iteration to run the simulation, $\mathcal{M}$ is the mesh of the simulation, $\mathcal{E}$ is the set of mesh entities, $\mathcal{D}$ is the set of computation domains used for computations, $\Delta$ is the set of data of the simulation, each one mapped onto a mesh entity, and finally $\Gamma$ is the set of computations. The set of computations $\Gamma$ is an ordered list of stencil and local kernels.
One can notice that this work, for now, is limited to a single mesh type for a given simulation.

### 2015-09-21 WOLFHPC15/article/conclusion.tex
corrections article reviewer2 + related + reduction taille -  Helene

Although MSL is a promising case study from DSLs to component-based runtimes, many works in progress aims at improving this first contribution. First, to more clearly show the improvement of DSLs maintainability using component-based back-end, an alternative \texttt{DDS} component is under study, using Global Arrays~\cite{Nieplocha:2006:AAP:1125980.1125985}. In addition to this, alternatives for the \texttt{Computations} component, computed by MSC, are under study such as a dump to an OpenMP~\cite{660313} code or the use of dynamic schedulers~\cite{Augonnet2011,Gautier:2013:XRS:2510661.2511383}. This last work on dynamic schedulers could also improve the expressivity of task parallelism in MSL, as discussed in Section~\ref{sect:eval}. Thus, it could bring interesting performance for the hybrid parallelization.

>>>>

Although MSL is a promising case study from DSLs to component-based runtimes, many works in progress aims at improving this first contribution. First, to more clearly show the improvement of DSLs maintainability using component-based back-end, an alternative \texttt{DDS} component is under study, using Global Arrays~\cite{Nieplocha:2006:AAP:1125980.1125985}. In addition to this, alternatives for the \texttt{Computations} component, computed by MSC, are under study such as a dump to a pure OpenMP~\cite{660313} code or the use of dynamic schedulers~\cite{Augonnet2011,Gautier:2013:XRS:2510661.2511383}. %This last work on dynamic schedulers could also improve the expressivity of task parallelism in MSL, as discussed in Section~\ref{sect:eval}. Thus, it could bring interesting performance for the hybrid parallelization.

### 2015-09-21 WOLFHPC15/article/evaluation.tex
corrections article reviewer2 + related + reduction taille -  Helene

The series-parallel tree decomposition $\Gamma_{tsp}$ of this simulation, extracted by the MSC transformation, is composed of 17~sequence nodes and 18~parallel nodes. Figure~\ref{fig:freq} represents for a given level of parallelism, \ie the number of tasks to perform concurrently, the number of time this level is observed in the final component assembly. One can notice that the level of task parallelism extracted from the Shallow water equations is limited by two sequential parts in the application (level 1). As a level of 16~parallel tasks is reached two times, and also five times for the fourth level, sequential restrictions could be amortized. However, it is rather clear that the task parallelization technique should be used to improve the data parallelism when reaching its limits, but not to use alone. Moreover, as the level of parallelism in the application is heterogenous, the number of threads to launch for task parallelism, and the number of cores to keep for data parallelism is difficult to choose.

@@

Figure~\ref{fig:exectime} illustrates the execution time for each step of the MSC transformation for an overall execution time of ten seconds. Execution times have been computed on a laptop with a bi-core Intel Core i5 1.4~GHz, and 8~GB of DDR3. 
One can notice that the transformation of $\Gamma_{dep}$ to a minimal series-parallel graph is the longest step of MSC, because of the removal of the forbidden N-shapes in the graph. Actually, the number of forbidden shapes removed in $\Gamma_{dep}$ is not counted by the algorithm, which applies a general solution (instead of finding each forbidden shape), but it seems that many of them appear.
%The \emph{N-shape}, represented in Figure~\ref{fig:n} is forbidden in a minimal series-parallel graph as it is not possible to exactly express it using sequences and parallel sections. However, a \emph{N-shape} could perfectly be handled by a dynamic scheduler, for example.
Thus, the fact that many \emph{N-shape} are removed in $\Gamma_{dep}$ shows that the creation of a static shedule of tasks may not be the best solution for complex simulations. It has to be noticed, however, that the all computation of $\Gamma_{tsp}$ is still usefull to detect how space loops can be merged in $\Gamma_{data}$.

\subsection{Preliminary Performance Evaluation}
To evaluate performance of the generated component-based parallel structure, we have proceeded as follows. Our current implementation of the MSCAC compiler generates a back-end code using the SkelGIS distributed data structure. For this reason, our evaluation compare the shallow water equations first parallelized with a pure SkelGIS code (data structures, applicators and interfaces of SkelGIS~\cite{CPE:CPE3494}), and second parallelized with MSL (which uses the SkelGIS data structure). Moreover, as SkelGIS library handles data parallelization for distributed memory architectures, we have limited for now our evaluations to the merged data parallelization of MSCAC (dumped from $\Gamma_{data}$).

Evaluations have been performed on the cluster \emph{Edel} of \emph{Grid'5000}\footnote{\url{https://www.grid5000.fr/}}. Figure~\ref{tab:hard} displays the features of this cluster. On \emph{Edel}, three different experiments have been performed that differ in the size of the Cartesian mesh and the number of time iterations: (1) $5,000 \times 5,000$, $500$ iterations; (2) $10,000 \times 10,000$, $500$ iterations; (3) $10,000 \times 10,000$, $2,000$ iterations.

\begin{figure}[!h]
\begin{center}
 \begin{tabular}{|c|c|c|}
    \hline
    Cluster & \textbf{Edel}\\
    \hline         
    Processor & 2 Intel Xeon E5520\\
    & (2.27 GHz)\\
    Cores/node & 8 \\
    Memory/node & 24 GB \\
    Compiler & OpenMPI [-O3] \\
    Network & InfiniBand 40G\\
    \hline
 \end{tabular}
   \caption{\label{tab:hard}Hardware of clusters used in the experiments.}
 \end{center}
\end{figure}

Figure~\ref{fig:perfs} reports the results of the three experiments. The first row of graphs is the number of iterations computed per second for various number of cores. The second row of graphs represents the execution time for various number of cores, using a logarithmic scale.

@@

One can notice that, for the three different experiments, execution times of MSL, which itself uses the SkelGIS distributed data structure, are improved compared to a pure SkelGIS parallelization. This result can be explained by the automatic generation of the parallel component-based runtime. Actually, in a pure SkelGIS code, communications between processes are hidden from the user through specific interfaces called \emph{applicators}~\cite{CPE:CPE3494}. Using MSL, on the other hand, communications are directly managed by $SYNC$ components, which are automatically placed in the component-based runtime by MSCAC. Thus, no overhead is introduced to hide communications.

In addition to this, the scalability slope, shown by the number of iterations per second, is close to SkelGIS. This is an important result as the execution time of MSL is at the same time improved compared to SkelGIS. Actually, the less the execution time is, the less the scalability capacity is. Thus, these results confirm the result of~\cite{l2c} that using a component-based runtime does not damage performance of the back-end code. Finally, we notice that SkelGIS has proved its scalability on the Shallow water equations compared to an MPI parallelization~\cite{CPE:CPE3494}, that makes our performance results relevant.

>>>>

The series-parallel tree decomposition $\Gamma_{tsp}$ of this simulation, extracted by the MSC transformation, is composed of 17~sequence nodes and 18~parallel nodes. Figure~\ref{fig:freq} represents for a given level of parallelism, \ie the number of tasks to perform concurrently, the number of time this level is observed in the final component assembly. One can notice that the level of task parallelism extracted from the Shallow water equations is limited by two sequential parts in the application (level 1). As a level of 16~parallel tasks is reached two times, and also five times for the fourth level, sequential restrictions could be amortized. Moreover, each identified task can itself be parallelized, introducing data parallelism on loops. Thus, it is rather clear that the task parallelization technique should be used to improve the data parallelism when reaching its limits, but not to use alone. Moreover, as the level of parallelism in the application is heterogenous, the number of threads to launch for task parallelism, and the number of cores to keep for data parallelism is difficult to choose.

@@

Figure~\ref{fig:exectime} illustrates the execution time for each step of the MSC transformation for an overall execution time of ten seconds. Execution times have been computed on a laptop with a bi-core Intel Core i5 1.4~GHz, and 8~GB of DDR3. 
One can notice that the transformation of $\Gamma_{dep}$ to a minimal series-parallel graph is the longest step of MSC, because of the removal of the forbidden N-shapes in the graph. %Actually, the number of forbidden shapes removed in $\Gamma_{dep}$ is not counted by the algorithm, which applies a general solution (instead of finding each forbidden shape), but it seems that many of them appear.
%The \emph{N-shape}, represented in Figure~\ref{fig:n} is forbidden in a minimal series-parallel graph as it is not possible to exactly express it using sequences and parallel sections. However, a \emph{N-shape} could perfectly be handled by a dynamic scheduler, for example.
Thus, the fact that many \emph{N-shape} are removed in $\Gamma_{dep}$ shows that the creation of a static shedule of tasks may not be the best solution for complex simulations. It has to be noticed, however, that the all computation of $\Gamma_{tsp}$ is still usefull to detect how space loops can be merged in $\Gamma_{data}$.

\subsection{Preliminary Performance Evaluation}
To evaluate performance of the generated component-based parallel structure, we have proceeded as follows. Our current implementation of the MSCAC compiler generates a back-end code using the SkelGIS distributed data structure as an underlying library. For this reason, our evaluation compare the shallow water equations first parallelized with a pure SkelGIS code (data structures, applicators and interfaces of SkelGIS~\cite{CPE:CPE3494}), and second parallelized with MSL (which uses the SkelGIS data structure only). Moreover, as SkelGIS library handles data parallelization for distributed memory architectures, we have limited for now our evaluations to the merged data parallelization of MSCAC (dumped from $\Gamma_{data}$).

Evaluations have been performed on the cluster \emph{Edel} of \emph{Grid'5000}\footnote{\url{https://www.grid5000.fr/}} for which each node is equiped with two 8-cores Intel Xeon E5520, 24GB of memory, and is interconnected with an Infiniband 40G network. %Figure~\ref{tab:hard} displays the features of this cluster. 
On \emph{Edel}, three different experiments have been performed that differ in the size of the Cartesian mesh and the number of time iterations: (1) $5,000 \times 5,000$, $500$ iterations; (2) $10,000 \times 10,000$, $500$ iterations; (3) $10,000 \times 10,000$, $2,000$ iterations.
% \begin{figure}[!h]
% \begin{center}
%  \begin{tabular}{|c|c|c|}
%     \hline
%     Cluster & \textbf{Edel}\\
%     \hline         
%     Processor & 2 Intel Xeon E5520\\
%     & (2.27 GHz)\\
%     Cores/node & 8 \\
%     Memory/node & 24 GB \\
%     Compiler & OpenMPI [-O3] \\
%     Network & InfiniBand 40G\\
%     \hline
%  \end{tabular}
%    \caption{\label{tab:hard}Hardware of clusters used in the experiments.}
%  \end{center}
% \end{figure}
Figure~\ref{fig:perfs} reports the results of the three experiments. The first row of graphs is the number of iterations computed per second for various number of cores. The second row of graphs represents the execution time for various number of cores, using a logarithmic scale.

@@

One can notice that, for the three different experiments, execution times of MSL, which itself uses the SkelGIS distributed data structure, are improved compared to a pure SkelGIS parallelization. This result can be explained by the automatic generation of the parallel component-based runtime. Actually, in a pure SkelGIS code, communications between processes are hidden from the user through specific interfaces called \emph{applicators}~\cite{CPE:CPE3494}. Using MSL, on the other hand, communications are directly managed by $SYNC$ components, which are automatically placed in the component-based runtime by MSCAC. Thus, no overhead is introduced to hide communications.
In addition to this, the scalability slope, shown by the number of iterations per second, is close to SkelGIS. This is an important result as the execution time of MSL is at the same time improved compared to SkelGIS. Actually, the less the execution time is, the less the scalability capacity is. Thus, these results confirm the result of~\cite{l2c} that using a component-based runtime does not damage performance of the back-end code. Finally, we notice that SkelGIS has proved its scalability on the Shallow water equations compared to an MPI parallelization~\cite{CPE:CPE3494}, that makes our performance results relevant.

### 2015-09-21 WOLFHPC15/article/introduction.tex
corrections article reviewer2 + related + reduction taille -  Helene

Domain Specific Languages (DSLs) are a promising solution to hide intricate details of HPC programming from non experts.
By being specific to a well-defined set of problems, those languages enable the generation of well parallelized and optimized code for various architectures while requiring only a limited amount of information from the developer; part of the knowledge is directly embedded in the compiler.
A critical question however concerns the specificity of DSLs.
Actually, the less specific the DSL is, the more information its users must provide, and the less automatic optimizations can be embedded in the compiler.
%On the other side of the scope
Conversely, the more specific the DSL is, the less applications can use it, and the more difficult it is to amortize the implementation cost of the DSL (language, compiler, and runtime).
Programming complexity, as well as non-maintainability and non-portability often seem to be simply deported from scientists to DSL designers.
% Moreover, one can consider that end users need as much DSLs as domains, which keeps the amount of work to write and maintain up-to-date DSLs out of hands.

@@

Many solutions to automatically optimize and parallelize numerical simulations have been proposed in the literature, either as specific frameworks~\cite{CPE:CPE3494,Trilinos-Overview,petsc-efficient} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation (such as fluid dynamics, magneto-hydrodynamics, molecular dynamics etc.) is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations. The DSL proposed in this paper, called \emph{Multi-Stencil Language} or MSL, is a DSL to generate a parallel component-based structure of a numerical simulation, that we call a \emph{multi-stencil program}. As discussed in the related work, this work is complementary to the optimization and parallelization of a single stencil kernel.

>>>>

Domain Specific Languages (DSLs) are a promising solution to hide intricate details of HPC programming from non experts.
By being specific to a well-defined set of problems, those languages enable the generation of well parallelized and optimized code for various architectures while requiring only a limited amount of information from the developer; part of the knowledge is directly embedded in the compiler.
A critical question however concerns the specificity of DSLs.
Actually, the less specific the DSL is, the more information its users must provide, and the less automatic optimizations can be embedded in the compiler.
%On the other side of the scope
Conversely, the more specific the DSL is, the less applications can use it, and the more difficult it is to amortize the implementation cost of the DSL (language, compiler, and runtime).
Programming complexity, as well as non-maintainability and non-portability often seem to be simply transferred from scientists to DSL designers.
% Moreover, one can consider that end users need as much DSLs as domains, which keeps the amount of work to write and maintain up-to-date DSLs out of hands.

@@

Many solutions to automatically optimize and parallelize numerical simulations have been proposed in the literature, either as specific frameworks~\cite{CPE:CPE3494,Trilinos-Overview,petsc-efficient} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation (such as fluid dynamics, magneto-hydrodynamics, molecular dynamics etc.) is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations~\cite{Ragan-Kelley:2013:HLC:2491956.2462176}. The DSL proposed in this paper, called \emph{Multi-Stencil Language} or MSL, is a DSL to generate a parallel component-based structure of a numerical simulation, that we call a \emph{multi-stencil program}. As discussed in the related work, this work is complementary to the optimization and parallelization of a single stencil kernel and is independent from implementation choices (as for example the form of the studied mesh).

### 2015-09-21 WOLFHPC15/article/ms_msc.tex
minor diff
corrections article reviewer2 + related + reduction taille -  Helene

\subsubsection*{Computation Domains and Their Dependencies}
A computation domain is a subset of a given kind of mesh entities, used for at least one computation.
For each computation domain, the kind of mesh entities on which it is mapped is specified (lines 3-4 and 12-14 on Figure~\ref{fig:grammar}).
By default, two computation domains of the same entities are handled as intersecting except when an explicit independence relation is mentioned in the \emph{independent} section of the language (lines 5-6 and 15-17 in Figure~\ref{fig:grammar}).
On lines 4 to 6 of Figure~\ref{fig:mslex}, computation domains are defined for each kind of mesh entities: \texttt{allcell}, \texttt{alledgex} and \texttt{alledgey}; they are used to represent the whole domain.
On lines 7 and 8, two other domains of \texttt{edgex} are defined: \texttt{part1edgex} and \texttt{part2edgex}; they are used to represent the boundaries of the domain.
On line 10 the independence relation between \texttt{part1edgex} and \texttt{part2edgex} is specified.

@@

\subsubsection*{Data parallelism optimizations: $\Gamma_{data}$}
The series-parallel tree decomposition $\Gamma_{tsp}$ is used to build the hybrid parallelization of the simulation, but it can also be used to optimize the data parallelization. In fact, it first seems natural to create the data parallel structure directly from $\Gamma_{sync}$, as the program is parallelized from data decomposition and applying the same sequential tasks on each processor. However, as a computation kernel writes a single data $w$ (to finely manage dependencies between tasks), the same computation domain might be iterated more than once, while some numerical expressions could be computed inside a joint space loop. For this reason, an optimized and \emph{fusionned} ordered list of computation $\Gamma_{data}$ is built from $\Gamma_{tsp}$. Actually, in $\Gamma_{tsp}$, computations of a parallel subtree with a same computation domain can be computed in the same space loop, in any order. Moreover, in $\Gamma_{tsp}$, two consecutive computations of a sequence subtree with a same computation domain can be computed in the same space loop, if the computation order of the sequence is kept. For example, from $\Gamma_{tsp}$ of Figure~\ref{fig:tsp}, as kernels $c_3$ and $c_5$ are executed inside a parallel section, and as their computation domains are the same (\texttt{alledgey}), a single component $K$ can be written peforming both computations in the same domain loop.

>>>>

\subsubsection*{Computation Domains and Their Dependencies}
A computation domain is a subset of a given kind of mesh entities, used for at least one computation.
For each computation domain, the kind of mesh entities on which it is mapped is specified (lines 3-4 and 12-14 on Figure~\ref{fig:grammar}).
By default, two computation domains of the same entities are handled as intersecting except when an explicit independence relation is mentioned in the \emph{independent} section of the language (lines 5-6 and 15-17 in Figure~\ref{fig:grammar}).
On lines 4 to 6 of Figure~\ref{fig:mslex}, computation domains are defined for each kind of mesh entities: \texttt{allcell}, \texttt{alledgex} and \texttt{alledgey}; they are used to represent the whole domain.
On lines 7 and 8, two other domains of \texttt{edgex} are defined: \texttt{part1edgex} and \texttt{part2edgex}; they could be used to represent the boundaries of the domain, for example, or any other sub-part of the whole domain. It is the user's responsability to define the needed domains for his simulation.
On line 10 the independence relation between \texttt{part1edgex} and \texttt{part2edgex} is specified.

@@

\subsubsection*{Data parallelism optimizations: $\Gamma_{data}$}
The series-parallel tree decomposition $\Gamma_{tsp}$ is used to build the hybrid parallelization of the simulation, but it can also be used to optimize the data parallelization. In fact, it first seems natural to create the data parallel structure directly from $\Gamma_{sync}$, as the program is parallelized from data decomposition and applying the same sequential tasks on each processor. However, as a computation kernel writes a single data $w$ (to finely manage dependencies between tasks), the same computation domain might be iterated more than once, while some numerical expressions could be computed inside a joint space loop. For this reason, an optimized and \emph{merged} ordered list of computation $\Gamma_{data}$ is built from $\Gamma_{tsp}$. Actually, in $\Gamma_{tsp}$, computations of a parallel subtree with a same computation domain can be computed in the same space loop, in any order. Moreover, in $\Gamma_{tsp}$, two consecutive computations of a sequence subtree with a same computation domain can be computed in the same space loop, if the computation order of the sequence is kept. For example, from $\Gamma_{tsp}$ of Figure~\ref{fig:tsp}, as kernels $c_3$ and $c_5$ are executed inside a parallel section, and as their computation domains are the same (\texttt{alledgey}), a single component $K$ can be written peforming both computations in the same domain loop.

### 2015-09-21 WOLFHPC15/article/related.tex
corrections article reviewer2 + related + reduction taille -  Helene

Many specific solutions exist to ease parallel programming of numerical simulations~\cite{Trilinos-Overview,petsc-efficient}. %Solutions such as PETSc~\cite{petsc-efficient} and Trilinos~\cite{Trilinos-Overview} are specific frameworks. They are made to solve many kinds of PDEs using well-known methods and solvers which are efficiently parallelized inside the library. Those frameworks also include solvers for implicit numerical schemes, not addressed in this work. However, using those solutions the user has to use and follow existing tools. In other words, it is not possible for a mathematician or a physician to write his own numerical method if it differs from existing ones. Moreover as PETSc and Trilinos are frameworks they do not propose a domain specific language to define computations. The purpose of those solutions are much more specific than our solution.
The closest solutions to MSL are DSLs which propose to generate parallel codes from a sequential imperative stencil description. For example, PATUS~\cite{citeulike12258902} and Pochoir~\cite{spaaTangCKLL11} propose a simple description to write stencil codes for general structured meshes, and OP2~\cite{Giles2011} and Liszt~\cite{DeVito2011LDS} for general unstructured meshes. Those solutions can be considered as stencil compilers. Actually they produce optimized (cache tiling, etc.) or parallel (CUDA, OpenCL, etc.) codes for a \emph{single} stencil computation.
Recently, domain specific languages for very specific numerical methods such as ExaSlang~\cite{Schmitt:2014:EDL:2691166.2691171} for multigrid solvers, and OPS~\cite{Reguly:2014:ODS:2691166.2691173} for meshes of meshes (hybrid mesh) have also been proposed. 
All those solutions let the user implement their own stencil codes in a sequential programming style, and produce high performance codes for shared memory architectures and sometimes distributed memory architectures. 
Most stencil compilers, though, handle the parallelization or the optimization of a single stencil kernel, considering that it represents the main computation time of numerical simulations. However, most real case numerical simulation are composed of more than one stencil kernel, involving one or more stencil shapes, and additional local kernels, as explained in Section~\ref{sect:multistencil}.

@@

The work presented in this paper takes place at a different abstraction level of numerical simulations and can be seen as complementary to stencil compilers or distributed data structures. Actually, the MSL language is agnostic and should work for any type of mesh and any neighborhood shape in stencil codes. The MSCAC compiler only needs this agnostic description of the overall simulation to generate a parallel pattern of the program. This parallel pattern then needs to be filled with implementation parameters and computation codes.
%. Thus, depending on the underlying distributed data structure used, those computation codes 
Computation codes could be generated by stencil compilers or could be written using interfaces of distributed data structures (as it is done using SkelGIS in this paper). As a result, MSL is not a new contribution to stencil compilers, or to distributed data structures, but proposes a coarse-grain automatic parallelization of an overall multi-stencil program.
Finally, MSL is a quite complex case study that indicates that component-based runtimes could improve code-reuse, separation of concerns and thus maintainability of DSLs without impacting performance.

>>>>

Many specific solutions exist to ease parallel programming of numerical simulations, %~\cite{Trilinos-Overview,petsc-efficient}. %Solutions such as PETSc~\cite{petsc-efficient} and Trilinos~\cite{Trilinos-Overview} are specific frameworks. They are made to solve many kinds of PDEs using well-known methods and solvers which are efficiently parallelized inside the library. Those frameworks also include solvers for implicit numerical schemes, not addressed in this work. However, using those solutions the user has to use and follow existing tools. In other words, it is not possible for a mathematician or a physician to write his own numerical method if it differs from existing ones. Moreover as PETSc and Trilinos are frameworks they do not propose a domain specific language to define computations. The purpose of those solutions are much more specific than our solution.
as for example DSLs which propose to generate parallel codes from a sequential imperative stencil description. For example, PATUS~\cite{citeulike12258902} and Pochoir~\cite{spaaTangCKLL11} propose a simple description to write stencil codes for general structured meshes, and OP2~\cite{Giles2011} and Liszt~\cite{DeVito2011LDS} for general unstructured meshes. Those solutions can be considered as stencil compilers. Actually they produce optimized (cache tiling, etc.) or parallel (CUDA, OpenCL, etc.) codes for a \emph{single} stencil computation.
Recently, domain specific languages for very specific numerical methods such as ExaSlang~\cite{Schmitt:2014:EDL:2691166.2691171} for multigrid solvers, and OPS~\cite{Reguly:2014:ODS:2691166.2691173} for meshes of meshes (hybrid mesh) have also been proposed. 
All those solutions let the user implement their own stencil codes in a sequential programming style, and produce high performance codes for shared memory architectures and sometimes distributed memory architectures. 
Most stencil compilers, though, handle the parallelization or the optimization of a single stencil kernel, considering that it represents the main computation time of numerical simulations. However, most real case numerical simulation are composed of more than one stencil kernel, involving one or more stencil shapes, and additional local kernels, as explained in Section~\ref{sect:multistencil}.

@@

Halide~\cite{Ragan-Kelley:2013:HLC:2491956.2462176} is a DSL which focusses on the automatic parallelization and optimization of stencil pipelines in image processing. Thus, this work is the closest to MSL. However, at least one important difference can be noticed: the specific domain it is applied on. Actually, image processing is applied onto 2D or 3D grids, while explicit numerical schemes to solve PDEs can be applied on many different kind of meshes (2D, 3D grids, but also unstructured or hybrid meshes). This main difference is the reason why MSL is a descriptive and agnostic language, and why its optimizations capabilities are limited compared to DSLs applied on known data structures (making available tiling optimizations etc.).

The work presented in this paper takes place at a different abstraction level of numerical simulations and can be seen as complementary to stencil compilers or distributed data structures. The MSCAC compiler only needs the agnostic description of the overall simulation to generate a parallel pattern of the program. This parallel pattern then needs to be filled with implementation parameters and computation codes.
%. Thus, depending on the underlying distributed data structure used, those computation codes 
Computation codes could be generated by stencil compilers or could be written using interfaces of distributed data structures (as it is done using SkelGIS in this paper). As a result, MSL is not a new contribution to stencil compilers, to tiling optimizations, or to distributed data structures, but proposes a coarse-grain automatic parallelization of an overall multi-stencil program.
Finally, MSL is a quite complex case study that indicates that component-based runtimes could improve code-reuse, separation of concerns and thus maintainability of DSLs without impacting performance.
